prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-06 inst/MyArticle/paper.Rmd
supp

The code snippets in this section demonstrate the basic usage of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. As input data we use Ionstar dataset.
Peptide intensities are $\log_2$ transformed and robust z-score scaled but then rescaled to preserves the original range of the data using the method `robscale`. We show the distribution of the normalized intensities in Figure \@ref(fig:prepro) Panel A. Protein intensities are estimated using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. The boxplot (Panel D) shows the distribution of the protein intensity estimate in each condition. Finally, Panel C shows the distribution of the protein standard deviations for each condition.

@@

(ref:exampleContrasts) Left panel - histogram of $p$-values. Right panel - volcano plots.

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 6}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()

@@

To ease the usage barriers of the R package to users not being so familiar with statistics and R programming, we integrated some workflows into our Local Laboratory Information Management System (LIMS) [b-fabric](https://fgcz-bfabric.uzh.ch/) @bfabric.
This integration enables our users to select the input and basic settings in a graphical user interface. As an output, the user receives a report including the R markdown file and all necessary input files to reprocess the analysis using R on their devices. In this way, we also support the reproducible science efforts of the community.

@@

The following code demonstrates how an analysis of MSFragger output can be performed using prolfqua.
First, a tidy table with protein abundances is generated by reading protein abundances reported by MSFragger in the combined_protein.tsv file.
The samples are annotated using the data stored in the annotation.xlsx file. Next, an _AnalysisTableAnnotation_ R6 object is created. 
Since bottom-up proteomics data is hierarchical a protein has peptides, which might be modified and create precursors. To describe this hierarchy, AnalysisTableAnnotation has a hierarchy field that stores a list with an entry for each hierarchy level. Since combined_portein.tsv only holds protein level data, the hierarchy list has only one element, and we need to specify which column does contain the protein identifiers.
We also need to specify which column contains the protein abundances we want to use for the data analysis.
Finally, we have to specify which columns contain the explanatory variables of the analysis. The AnalysisTableAnnotation has the field factors, a list with as many entries as explanatory variables. Here we have only one explanatory variable, the dilution, specified in the column 'sample.'

@@

Here we compare the benchmark results for the `r BiocStyle::Biocpkg("proDA")` model and the linear model with variance moderation, but using only the protein where inference could be performed for both models. We see that the fold change estimates obtained with the `r BiocStyle::Biocpkg("proDA")` package show better diagnostic accuracy. Figure \@ref(fig:proDAAllBenchmarks), compares the fold change estimates obtained using both methods. However, the $t-statistics$ or $p-values$ show slightly worse diagnostic accuracy.

```{r proDABenchmarks}
prodB <- allBenchmarks$benchmark_proDA$clone(deep=TRUE)
proda <- prodB$.data
modB <- allBenchmarks$benchmark_ProtModerated

mod <- modB$.data
mod <- mod[!is.na(mod$statistic),]
modSum <- modB$pAUC_summaries()

prodaMod <- dplyr::semi_join(proda, mod, by = c("contrast", "protein_Id"))
prodB$.data <- prodaMod
d <- prodB$missing_contrasts()
prodSum <- prodB$pAUC_summaries()

@@

```{r proDAAllBenchmarks, fig.cap="Comparison of fold change estimates produced by proDA and using the missing data modelling method implemented in prolfqua."}
proda <- allBenchmarks$benchmark_proDA$data()
imp <- allBenchmarks$benchmark_imputation$data()
imp <- allBenchmarks$benchmark_ProtModerated$data()

x <- dplyr::inner_join(imp, proda, by = c("contrast","protein_Id", "TP", "species"), suffix = c(".imp",".proda"))
ggplot2::ggplot(x,ggplot2::aes( x = diff.imp, y = diff.proda) ) + ggplot2::geom_point() + ggplot2::facet_wrap(~contrast + species) + ggplot2::geom_hline(yintercept = 0, color="red") + ggplot2::geom_vline(xintercept = 0, color="red")

@@

(ref:tabCompletion) The screenshot displays the command-line completion (tab completion) of RStudio on the `prolfqua::LFQData` R6 object. In the example, it shows the getter methods of the object.

```{r tabCompletion, echo=FALSE, fig.cap="(ref:tabCompletion)", out.width = '66%'}
knitr::include_graphics("codeSnippet1TabCompletion.png")
```

>>>>

The code snippets in this section demonstrate the basic usage of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. As input data we use subset of the Ionstar dataset containing $163$ proteins and $1258$ peptides. Peptide intensities are $\log_2$ transformed and robust z-score scaled but then rescaled to preserves the original range of the data using the method `robscale`. We show the distribution of the normalized intensities in Figure \@ref(fig:prepro) Panel A. Protein intensities are estimated using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. The boxplot (Panel D) shows the distribution of the protein intensity estimate in each condition. Finally, Panel C shows the distribution of the protein standard deviations for each condition.

@@

(ref:exampleContrasts) Panel A - Histogram showing the distribution of p-values for $163$ proteins. Panel B -  Volcano plot showing $-\log_{10}$ transformed FDR as function of the difference between groups for $163$ proteins.

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
# specify differences among conditions
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()

@@

To ease the usage barriers of the R package to users not being so familiar with statistics and R programming, we integrated some workflows into our Local Laboratory Information Management System (LIMS) [b-fabric](https://fgcz-bfabric.uzh.ch/) @bfabric.
This integration enables our users to select the input and basic settings in a graphical user interface. Then, as an output, the user receives a report including the R markdown file and all necessary input files to reprocess the analysis using R on their devices. In this way, we also support the reproducible science efforts of the community.

@@

The following code demonstrates how an analysis of MSFragger output can be performed using prolfqua.
First, a tidy table with protein abundances is generated by reading protein abundances reported by MSFragger in the combined_protein.tsv file.
The samples are annotated using the data stored in the annotation.xlsx file. Next, an _AnalysisTableAnnotation_ R6 object is created. 
Since bottom-up proteomics data is hierarchical a protein has peptides, which might be modified and create precursors. To describe this hierarchy, AnalysisTableAnnotation has a hierarchy field that stores a list with an entry for each hierarchy level. Since combined_portein.tsv only holds protein level data, the hierarchy list has only one element, and we need to specify which column does contain the protein identifiers.
We also need to specify which column contains the protein abundances we want to use for the data analysis.
Finally, we have to specify which columns contain the explanatory variables of the analysis. The AnalysisTableAnnotation has the field factors, a list with as many entries as there are explanatory variables. Here we have included two explanatory variable, the dilution, specified in the column 'sample', and 'run' in the column 'run_ID' representing the order of the measurement.

@@

Here we compare the benchmark results for the `r BiocStyle::Biocpkg("proDA")` model and the linear model with variance moderation. We see that the fold change estimates obtained with the `r BiocStyle::Biocpkg("proDA")` package show better diagnostic accuracy.  However, the $t-statistics$ or $p-values$ show slightly worse diagnostic accuracy (Figure \@ref(fig:proDABenchmarks). Figure \@ref(fig:proDAAllBenchmarks), compares the fold change estimates obtained using both methods.

(ref:proDABenchmarks) The barplots show the difference of the $pAUC_10$ obtained when using the proDA package or when using prolfqua. Postitive values indicate a better diagnostic performance of proDA while negative values a better diagnostic performance of prolfqua.

```{r proDABenchmarks, fig.cap="(ref:proDABenchmarks)"}
prodB <- allBenchmarks$benchmark_proDA$clone(deep=TRUE)
proda <- prodB$.data
modB <- allBenchmarks$benchmark_ProtModerated

mod <- modB$.data
mod <- mod[!is.na(mod$statistic),]
modSum <- modB$pAUC_summaries()
prodaMod <- dplyr::semi_join(proda, mod, by = c("contrast", "protein_Id"))
prodB$.data <- prodaMod

d <- prodB$missing_contrasts()
prodSum <- prodB$pAUC_summaries()

@@

```{r proDAAllBenchmarks, fig.cap="Comparison of fold change estimates produced by proDA and prolfqua simple impute method."}
proda <- allBenchmarks$benchmark_proDA$data()
imp <- allBenchmarks$benchmark_imputation$data()
imp <- allBenchmarks$benchmark_ProtModerated$data()

x <- dplyr::inner_join(imp, proda, by = c("contrast","protein_Id", "TP", "species"), suffix = c(".imp",".proda"))
ggplot2::ggplot(x,ggplot2::aes( x = diff.imp, y = diff.proda) ) + ggplot2::geom_point() + ggplot2::facet_wrap(~contrast + species) + ggplot2::geom_hline(yintercept = 0, color="lightgray") + ggplot2::geom_vline(xintercept = 0, color="lightgray") + ggplot2::geom_abline(slope = 1) + ggplot2::theme_minimal()

@@
